The influence of small reservoirs on hydrological drought propagation in space and time

### 2022-10-26 LateX/Feb-3-22-latex-templates/paper_Colombo.tex
images, tables, biblio, appendix

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{bm}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

@@

%% ------------------------------------------------------------------------ %%
%  Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

% Example: \title{This is a test title}

@@

Droughts are defined as periods of exceptional lack of water that negatively impact human activities or environmental demands. (Van Loon et al., 2016). When drought hits, the most common effect is water shortages, which can lead to food shortages if agriculture is affected. Additional drought impacts include drying rivers and lakes, water use restrictions, reduced electricity production, dying forests, and wildfires. Combined these can cause famine, diseases, and migration.  Human activities can both aggravate and alleviate hydrological droughts, but the former case has been found more dominant (Van Loon et al., 2022). Identifying droughts’ drivers, both natural and human, is a major scientific challenge (Walker et al., 2022; Zaniolo et al., 2018). Thus, dealing with droughts also requires identifying how humans induce and modify exposure and vulnerability to drought.\\
Articulating the role of drought preparedness in the context of watershed management areas with a long term view can be an important practice to mitigate drought events (Gutiérrez et al., 2014). Reservoirs are built in dryland regions both as a drought preparedness measure and as response to the growing water demand (Rabelo et al., 2021). However, reservoir construction without a holistic watershed management approach can in turn generate higher water demand, and at the same time reduces the incentive for adaptive actions on other levels. Therefore, reservoirs can intensify or even induce drought events, since storing water upstream worsened hydrological drought downstream (Van Langen et al., 2021). This, in turn, can result in pressure to build more reservoirs, which may further aggravate the problem (Di Baldassarre et al., 2018). Both large publicly managed reservoirs and smaller privately owned reservoirs can play a role in this process.\\
Many places across the world show forms of dense networks of small reservoirs: Australia (Fowler et al., 2015), France (Habets et al., 2014), Ghana (Annor et al., 2009), North-East Brazil (Mamede et al., 2012), South India (Mialhe, 2008), and Syria (Avisse et al., 2017), for example. The influence of small reservoirs on water distribution and droughts is still not fully understood. They are commonly built without regulation nor monitoring in regions where there is a general lack of observational data. This generates uncertainty on their number and capacity, increasing the difficulty of their analysis (Habets et al., 2018). Small reservoirs, when analyzed individually, are not expected to cause major impacts to a hydrological system, since their maximum storage capacity is some orders of magnitude lower than strategic reservoirs (medium and large-sized reservoirs located on main rivers at the sub-basin’s outlet). However, the accumulated effect of a network of small reservoirs can lead to a 30\% increase in the duration of hydrological droughts (Ribeiro Neto et al., 2022). The investigation of the hydrological impact in time and space of a Dense Reservoir Network (DRN) is then highly relevant, both from a community perspective and from a drought and water management one. Moreover, diagnosing droughts also needs the evaluation of the drought management measures in place (Walker et al., 2022).\\
The impact of reservoirs on drought development has been already researched. However, most of these studies focused on large and medium-sized reservoirs (e.g. van Langen et al., 2021). Although some studies have analyzed the effect of small reservoirs on sediment/water dynamics (e.g. Mamede et al, 2018), the hydrological availability (e.g. Krol et al. 2011), evaporation losses, and streamflow impact (e.g. Wisser et al., 2010, Malveira et al.2012), there is a scientific gap related to the cumulative effect of a DRN on drought in a complex hydrological system (Walker et al., 2022). However, direct consideration of the DRN effects on droughts have started to be considered, such as in Ribeiro Neto et al. (2022), where a novel analysis on this topic is proposed and tested in the Riacho do Sangue watershed in North-East Brazil.Integrating DRN's role in a broader water management perspective which does not only consider strategic centralized reservoirs may help answer one of the “unsolved problems in hydrology”, the number 22: “What are the synergies and trade-offs between societal goals related to water management?” (Blöschl et al., 2019).\\
The aim of this research is to evaluate the spatio-temporal influence of a dense network of (small) reservoirs on drought evolution and its impacts on water availability at catchment scale. At the same time, assess the network modeling in a complex basin. We applied a methodology based on the combination of semi-distributed hydrological modeling with a framework for monitoring the evolution of drought events. This methodology was tested for the Banabuiú watershed in the state of Ceara in Brazil.

\section{Materials and Methods}
To assess the impacts on drought evolution and magnitude generated by the DRN presence, differences between case study watershed scenarios, including and excluding the DRN, have been analyzed (2.3). Modeling the Banabuiu basin has been selected as the way to represent these scenarios (2.2). The small reservoirs capacities have been estimated to be able to model the network (2.1).
\subsection{Study area}
%Figure 1: Study area framing. North-East Brazil, Ceará and Banabuiú catchment on the left. On the right, centralized reservoirs location and storage capacity, overlayed to DRN’s density in the Banabuiú basin.
The semi-arid region of Brazil (North-East Brazil) occupies about 11\% of the Brazilian territory (1,542,000 $km^2$) and contains 30\% of its population (Marengo et al., 2020). It is a representative example of a region with a high concentration of reservoirs with great variation in size and storage capacity. North-East Brazil (NEB) has a highly irregular spatio-temporal precipitation regime, which means the region is frequently affected by intense drought events. Sixteen out of the last 25 years registered rainfall below normal in the region. In 2010 the most severe drought ever recorded started, which ended in 2018 (Marengo et al. 2016, 2018). The combination of the predominance of soils and geology with low water storage capacity produces a dependency on the superficial storage of water (Marengo, Alves, et al., 2017; Marengo, Torres, & Alves, 2017). Water supply then relies on the reservoirs that regularize discharges. Their storage capacity varies from small reservoirs used on private properties to large reservoirs used for urban supply, industrial demands, and large irrigation areas (also called hydrosystems). A total of 17,083 reservoirs with surface area greater than 5 ha are located in the NEB. This system has a cumulative capacity of 707.36 billion $m^3$, according to the database from Reservoir Monitoring System of the Brazilian National Water Agency (Agência Nacional de Águas e Saneamento Básico) in 2016 (Nascimento and Ribeiro Neto, 2017). A representative example of this situation is Ceará, a 148,886 km² wide state in semi-arid Brazil, where it is estimated the presence of 105,813 reservoirs with more than 20 meters in length (FUNCEME, 2021). Ceará presents areas with reservoir concentration higher than 7 reservoirs per $km^2$ (Ribeiro Neto et al., 2022), which greatly exceed other high concentration areas such as India (4.2 reservoirs/$km^2$) and Australia (6.1 reservoirs/$km^2$) (Rabelo et. al, 2021).\\
The study area is the Banabuiu river watershed located in the state of Ceará which covers an area of approximately 19,000 $km^2$. The river network, starting from the main tributary Rio Quixeramobim, delineates the division of the basin into sub-basins and determines the distribution of the strategic reservoirs, placed along the network branches. The Banabuiú basin counts 19 monitored reservoirs with storage capacity above 1·$Hm^3$ ($10^6 m^3$) , with the biggest being Arrojado Lisboa at 1600 $Hm^3$. These reservoirs, shown in Figure 1, are continuously monitored and managed by the State water agency (Water Resources Company, COGERH) and account for 2790 $Hm^3$, 93\% of the overall water storage capacity in the basin. At the same time, a cumulative 210 $Hm^3$ storage capacity is retained in approximately 10,000 small reservoirs, which constitute the DRN and are visualized together with strategic reservoirs present in the basin in Figure 1. Despite their small proportion in relation to total capacity (7.1\%), small reservoirs have an important role in the water supply of small rural communities, whose needs are not met by the large hydrosystems, as the water stored in small reservoirs enables local distribution. In the region, most of the small reservoirs aim to meet the water necessities of subsistence farmers only in the short term (6-8 months in the dry season). The small reservoirs also serve as sediment detention basins, retaining a considerable amount of sediment generated within the catchment and extending the life-time of larger ones located downstream (MAMEDE et al. 2018). However, their presence can reduce the potential yield of strategic reservoirs which may determine a lower possibility to store and distribute water during droughts (Krol et al., 2011).

\subsection{DRN capacity estimation}
%Table 1. Repartition of the reservoirs composing the Dense Reservoir Network in the Banabuiu basin.
The capacity of the DRN-component reservoirs is needed to represent the network in the hydrological model.  Neither capacity nor surface area were available. Therefore, a capacity estimation has been obtained starting from the Joint Research Center’s (JRC) Global Surface Water Explorer, which provides a worldwide database of surface water imagery (Pekel et al., 2016). The maximum water extent raster, which maps the maximum extent of water bodies in an area, was connected with the locations of the DRN (surveyed in FUNCEME et al. 2021), associating each reservoir location to the nearest water body through a nearest neighbor algorithm. In this way, 10,355 small reservoirs had their maximum area associated. Reservoirs without a corresponding water body in JRC’s representation (approximately 7,000) were left out of the final configuration, represented in Figure 1 and summarized in Table 1. The causes of them not being represented may lay in their small size, which is difficult to be detected and validated through JRC’s satellite images, or in inaccuracies in the survey operated by FUNCEME et al. (2021), where small topographic depressions could have been erroneously interpreted as reservoirs, such as natural depressions in lowlands and depressions close to roads. The capacity of each reservoir (V) was then obtained from the area (A) through Molle’s equation (Equation 1), where α and K parameters are equal to average values of 2.7 and 1,500 respectively, as used in literature in North-East Brazil (Mamede et al., 2018, 2012; Molle 1994). In this study, a reservoir is considered small when its capacity is lower or equal to 500,000 $m^3$ (0.5 $Hm^3$), since this was the highest storage capacity of the unmonitored reservoirs in the basin.
%\begin{linenomath*}
%\begin{equation}
%	V = K \cdot (\frac{A}{ \alfa \cdot K})^\frac{ \alfa }{ \alfa - 1 }
%\end{equation}
%\end{linenomath*}

\subsection{Modeling approach}
\subsubsection{WASA-SED model}
The model selected to model the Banabuiú river basin is WASA-SED (Water Availability in Semi-Arid environments-SEDiments), which has been developed within the joint Spanish-Brazilian-German research project SESAM (Sediment Export from Semi-Arid Catchments: Measurement and Modelling) and utilized to simulate North-East Brazil’s conditions (Güntner, 2002; Güntner et al., 2004; Mueller et al., 2010). The model simulates the runoff and erosion processes at meso-scale, for domains ranging from several hundreds to thousands of square kilometers. It uses a hierarchical top-down disaggregation scheme, dividing each sub-basin of the model into landscape units (based on the Soil and Terrain Digital Database (SOTER) concept (Oldeman and van Engelen, 1993)), represented by multiple terrain components (defined by slope-gradient, length, soil and soil-vegetation components). The hydrological module is fully described in Güntner (2002) and Güntner et al. (2004), implementing for example equations to account for interception losses, evaporation and transpiration (modified Penman-Monteith approach (Shuttleworth and Wallace, 1985)) and for infiltration (Green-Ampt approach (Green and Ampt, 1911)) and other soil and vegetation related processes. The model can simulate both medium and large  centralized strategic reservoirs and small diffuse networks of reservoirs (Güntner et al., 2004). The user manual, the link to the source code and other useful sources can be found at \url{github.com/TillF/WASA-SED}. Detailed WASA-SED model parameterization for Banabuiú river basin was carried out by Costa et al. (2013), and adopted in this work.

\subsubsection{Small reservoirs representation}
The modeling approach to the reservoirs needs to classify them into strategic and small reservoirs, according to location and size (Güntner et al., 2004). The strategic reservoirs listed in Table 1 are individually parameterized in WASA-SED, also providing their daily regulated discharge time series. Their water balance is calculated explicitly and individually for each reservoir. Due to the uncontrolled and unmonitored nature of small reservoirs composing the DRN, not enough information is available concerning dam building, location, size and water use. They are grouped into 5 storage capacity classes and the water balance is computed for a hypothetical reservoir with mean characteristics, representative of each class and sub-basin. The water storage volume is then given by the product of the water storage of the representative reservoir by the total number of reservoirs from that class within a sub-basin. The generated runoff of the sub-basin is distributed among the reservoir classes through a cascade routing scheme starting from the lowest class to the highest, using a weighting factor computed as a ratio between the runoff contributing area of that reservoir class and the sub-basin area (Mamede et al., 2018). A detailed description of the calculation of water fluxes through the classes of reservoirs can be found in Mamede (2008), while equations and other information about the hydrological and reservoir module can be found in Güntner (2002).

\subsubsection{Calibration and validation}
Calibration was focused on the “scaling factor” parameter, which modifies the soil saturated hydraulic conductivity, which is the most sensitive parameter  (Güntner et al., 2002; Güntner et al., 2021). The calibration was carried out automatically following the sub-basins routing path of the model from upstream to downstream, testing a set of 35 values on each sub-basin (ranging from 0.2 to 7). For each run, five model performance indexes were computed (R2, NSE, KGE, PBIAS and NRMSE) comparing the modeled volume time series for the sub-basin’s strategic reservoir (or the direct downstream one in case the sub-basin did not present one itself) with the observed one. The performance indexes were selected based on recent usage in hydrological modeling (Marahatta et al., 2021; Knoben et al., 2019; Uniyal et al., 2019). The parameter that would produce the best performance was selected. The model was calibrated with 70\% of the series, running it from 1980 to 2006, it was then validated on the remaining 30\% of the series, from 2007 to 2018. Wet, normal and dry years were well represented in both the selected periods (calibration and validation) of the gauges. Therefore, the high inter-annual streamflow variability (CV greater than 1) was taken into account for both calibration and validation periods.

\subsubsection{Scenarios generated and used in the analyses}
%Figure 2: Visualization of the modeled scenarios design: Large Reservoirs (LR), centralized reservoirs only; All Reservoirs (AR), centralized reservoirs and DRN; Naturalized (N), Arrojado Lisboa only; and Small Reservoirs (SR), DRN and Arrojado Lisboa only.
%Substitute all Figure 4 references to references to this figure (adding tag{fig:scenario} or something similar)
Four different configurations were modeled, as visualized in Figure 4. The All Reservoirs (AR) configuration represents the real condition, with the co-existence of the small and large reservoirs. The dense reservoir network was removed to keep only large reservoirs in the Large Reservoirs (LR) scenario. These were also removed to obtain two scenarios not influenced by them: Small Reservoirs (SR), representing only small reservoirs and Arrojado Lisboa, and Naturalized (N), representing only Arrojado Lisboa. The only reservoir present in all the scenarios is Arrojado Lisboa, the most downstream, used as a reference for the comparisons in all the analyses. The differences between AR and LR scenarios have been explored in order to assess the effects of the DRN on droughts’ evolution in time and space. N and SR scenarios have been useful to explore the DRN effects without the influence of large reservoirs.

\subsection{Drought analyses}
To effectively consider the effects that the existence of a DRN produces on drought evolution, two domains have been considered: time, to address impacts on frequency and duration of drought, and space, to include the effects on water distribution. This dual aspect has been analyzed through the Drought Cycle Analysis (2.4.1) (Ribeiro Neto et al., 2022) and the Downstreamness Analysis (2.4.2) (van Oel et al., 2018). The different scenarios created through the WASA-SED model will be compared with these methods in order to estimate the influence of the DRN on hydrological drought evolution and propagation.

>>>>

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\usepackage{bm}

\usepackage{array}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

@@

@@

Droughts are defined as periods of exceptional lack of water that negatively impact human activities or environmental demands \cite{VanLoon2016}. When drought hits, the most common effect is water shortages, which can lead to food shortages if agriculture is affected. Additional drought impacts include drying rivers and lakes, water use restrictions, reduced electricity production, dying forests, and wildfires. Combined these can cause famine, diseases, and migration.  Human activities can both aggravate and alleviate hydrological droughts, but the former case has been found more dominant \cite{VanLoon2022}. Identifying droughts’ drivers, both natural and human, is a major scientific challenge \cite{Walker2022,Zaniolo2018}. Thus, dealing with droughts also requires identifying how humans induce and modify exposure and vulnerability to drought.\\
Articulating the role of drought preparedness in the context of watershed management areas with a long term view can be an important practice to mitigate drought events \cite{Gutierrez2014}. Reservoirs are built in dryland regions both as a drought preparedness measure and as response to the growing water demand \cite{Rabelo2021}. However, reservoir construction without a holistic watershed management approach can in turn generate higher water demand, and at the same time reduces the incentive for adaptive actions on other levels. Therefore, reservoirs can intensify or even induce drought events, since storing water upstream worsened hydrological drought downstream \cite{HvanLangen2021}. This, in turn, can result in pressure to build more reservoirs, which may further aggravate the problem \cite{DiBaldassarre2018}. Both large publicly managed reservoirs and smaller privately owned reservoirs can play a role in this process.\\
Many places across the world show forms of dense networks of small reservoirs: Australia \cite{Fowler2016}, France \cite{Habets2014}, Ghana \cite{Annor2009}, North-East Brazil \cite{Mamede2012}, South India \cite{Mialhe2008}, and Syria \cite{Avisse2017}, for example. The influence of small reservoirs on water distribution and droughts is still not fully understood. They are commonly built without regulation nor monitoring in regions where there is a general lack of observational data. This generates uncertainty on their number and capacity, increasing the difficulty of their analysis \cite{Habets2018}. Small reservoirs, when analyzed individually, are not expected to cause major impacts to a hydrological system, since their maximum storage capacity is some orders of magnitude lower than strategic reservoirs (medium and large-sized reservoirs located on main rivers at the sub-basin’s outlet). However, the accumulated effect of a network of small reservoirs can lead to a 30\% increase in the duration of hydrological droughts \cite{RibeiroNeto2022}. The investigation of the hydrological impact in time and space of a Dense Reservoir Network (DRN) is then highly relevant, both from a community perspective and from a drought and water management one. Moreover, diagnosing droughts also needs the evaluation of the drought management measures in place \cite{Walker2022}.\\
The impact of reservoirs on drought development has been already researched. However, most of these studies focused on large and medium-sized reservoirs \cite <e.g.,>{HvanLangen2021}. Although some studies have analyzed the effect of small reservoirs on sediment/water dynamics  \cite <e.g.,>{Mamede2018}, the hydrological availability \cite <e.g.,>{Krol2011}, evaporation losses, and streamflow impact \cite <e.g.,>{Wisser2010, Malveira2012}, there is a scientific gap related to the cumulative effect of a DRN on drought in a complex hydrological system \cite{Walker2022}. However, direct consideration of the DRN effects on droughts have started to be considered, as in \citeA{RibeiroNeto2022}, where a novel analysis on this topic is proposed and tested in the Riacho do Sangue watershed in North-East Brazil.Integrating DRN's role in a broader water management perspective which does not only consider strategic centralized reservoirs may help answer one of the “unsolved problems in hydrology”, the number 22: “What are the synergies and trade-offs between societal goals related to water management?” \cite{Bloschl2019}.\\
The aim of this research is to evaluate the spatio-temporal influence of a dense network of (small) reservoirs on drought evolution and its impacts on water availability at catchment scale. At the same time, assess the network modeling in a complex basin. We applied a methodology based on the combination of semi-distributed hydrological modeling with a framework for monitoring the evolution of drought events. This methodology was tested for the Banabuiú watershed in the state of Ceará in Brazil.

\section{Materials and Methods}
To assess the impacts on drought evolution and magnitude generated by the DRN presence, differences between case study watershed scenarios, including and excluding the DRN, have been analyzed (\ref{sec:analyses}). Modeling the Banabuiú basin has been selected as the way to represent these scenarios (\ref{sec:model}). The small reservoirs capacities have been estimated to be able to model the network (\ref{sec:drncap}).

\subsection{Study area}
%Figure 1
\begin{figure}
 \noindent\includegraphics[width=\textwidth]{images/Figure_1.png}
 \caption{Study area framing. North-East Brazil, Ceará and Banabuiú catchment on the left. On the right, centralized reservoirs location and storage capacity, overlayed to DRN’s density in the Banabuiú basin.}
 \label{fig1}
\end{figure}

The semi-arid region of Brazil (North-East Brazil) occupies about 11\% of the Brazilian territory (1,006,654 $km^2$) and has a population of 26 million inhabitants \cite{Marengo2020}. It is a representative example of a region with a high concentration of reservoirs with great variation in size and storage capacity. North-East Brazil (NEB) has a highly irregular spatio-temporal precipitation regime, which means the region is frequently affected by intense drought events. Sixteen out of the last 25 years registered rainfall below normal in the region. In 2010 the most severe drought ever recorded started, which ended in 2018 \cite{MarengoOrsini2018,Marengo2017}. The combination of the predominance of soils and geology with low water storage capacity produces a dependency on the superficial storage of water \cite{Marengo2017,Rossato2017}. Water supply then relies on the reservoirs that regularize discharges. Their storage capacity varies from small reservoirs used on private properties to large reservoirs used for urban supply, industrial demands, and large irrigation areas (also called hydrosystems). A total of 17,083 reservoirs with surface area greater than 5 ha are located in the NEB. This system has a cumulative capacity of 707.36 billion $m^3$, according to the database from Reservoir Monitoring System of the Brazilian National Water Agency (Agência Nacional de Águas e Saneamento Básico) in 2016 \cite{Nascimento2017}. A representative example of this situation is Ceará, a 148,886 km² wide state in semi-arid Brazil, where it is estimated the presence of 105,813 reservoirs with more than 20 meters in length \cite{FUNCEME2021}. Ceará presents areas with reservoir concentration higher than 7 reservoirs per $km^2$ \cite{RibeiroNeto2022}, which greatly exceed other high concentration areas such as India (4.2 reservoirs/$km^2$) and Australia (6.1 reservoirs/$km^2$) \cite{Rabelo2021}.\\
The study area is the Banabuiú river watershed located in the state of Ceará which covers an area of approximately 19,000 $km^2$. The river network, starting from the main tributary Rio Quixeramobim, delineates the division of the basin into sub-basins and determines the distribution of the strategic reservoirs, placed along the network branches. The Banabuiú basin counts 19 monitored reservoirs with storage capacity above 1·$Hm^3$ ($10^6 m^3$) , with the biggest being Arrojado Lisboa at 1600 $Hm^3$. These reservoirs, shown in Figure~\ref{fig1}, are continuously monitored and managed by the State water agency (Water Resources Company, COGERH) and account for 2790 $Hm^3$, 93\% of the overall water storage capacity in the basin. At the same time, a cumulative 210 $Hm^3$ storage capacity is retained in approximately 10,000 small reservoirs, which constitute the DRN and are visualized together with strategic reservoirs present in the basin in Figure~\ref{fig1}. Despite their small proportion in relation to total capacity (7.1\%), small reservoirs have an important role in the water supply of small rural communities, whose needs are not met by the large hydrosystems, as the water stored in small reservoirs enables local distribution. In the region, most of the small reservoirs aim to meet the water necessities of subsistence farmers only in the short term (6-8 months in the dry season). The small reservoirs also serve as sediment detention basins, retaining a considerable amount of sediment generated within the catchment and extending the life-time of larger ones located downstream \cite{Mamede2018}. However, their presence can reduce the potential yield of strategic reservoirs which may determine a lower possibility to store and distribute water during droughts \cite{Krol2011}.

\subsection{DRN capacity estimation}\label{sec:drncap}

%Table 1
\begin{table}
 \caption{Repartition of the reservoirs composing the Dense Reservoir Network in the Banabuiú basin.}\label{tab1}
 \centering
 \begin{tabular*}{\textwidth}{m{1.5cm} m{2cm} m{2cm} m{2cm} m{2.3cm} m{2.3cm}}
 \hline
  Reservoir class & Storage volume - upper limit &  Number of reservoirs & Percentage of reservoirs & Class cumulative storage volume & Class cumulative storage volume\\
  & [$m^3$] & & [\%] & [$Hm^3$] & [\%]\\
 \hline
1	& 5000	 & 6977	&	67.4	 &	7.2  &	3.4\\
2	& 25000	 & 1787	&	17.3	 &	21	 & 10\\
3	& 50000	 & 546	&	5.3	 &	20	 & 9.4\\
4	& 100000	 & 447	&	4.3	 &	32	 & 14.8\\
5	& 500621	 & 598	&	5.8	 &	130	 & 62.4\\
 \hline
Total &	-	& 10355 &	100	 &  210.2 &	100\\
 \end{tabular*}
\end{table}

The capacity of the DRN-component reservoirs is needed to represent the network in the hydrological model.  Neither capacity nor surface area were available. Therefore, a capacity estimation has been obtained starting from the Joint Research Center’s (JRC) Global Surface Water Explorer, which provides a worldwide database of surface water imagery \cite{Pekel2016}. The maximum water extent raster, which maps the maximum extent of water bodies in an area, was connected with the locations of the DRN (surveyed in \citeA{FUNCEME2021}), associating each reservoir location to the nearest water body through a nearest neighbor algorithm. In this way, 10,355 small reservoirs had their maximum area associated. Reservoirs without a corresponding water body in JRC’s representation (approximately 7,000) were left out of the final configuration, represented in Figure~\ref{fig1} and summarized in Table~\ref{tab1}. The causes of them not being represented may lay in their small size, which is difficult to be detected and validated through JRC’s satellite images, or in inaccuracies in the survey operated by \citeA{FUNCEME2021}, where small topographic depressions could have been erroneously interpreted as reservoirs, such as natural depressions in lowlands and depressions close to roads. The capacity of each reservoir (V) was then obtained from the area (A) through Molle’s equation (Equation 1), where $\alpha$ and K parameters are equal to average values of 2.7 and 1,500 respectively, as used in literature in North-East Brazil \cite{Mamede2018,Mamede2012,Molle1994}. In this study, a reservoir is considered small when its capacity is lower or equal to 500,000 $m^3$ (0.5 $Hm^3$), since this was the highest storage capacity of the unmonitored reservoirs in the basin.
\begin{linenomath*}
	\begin{equation}
	V = K \cdot (\frac{A}{\alpha\cdot K})^\frac{\alpha}{\alpha -1}
	\end{equation}
\end{linenomath*}

\subsection{Modeling approach}\label{sec:model}
\subsubsection{WASA-SED model}
The model selected to model the Banabuiú river basin is WASA-SED (Water Availability in Semi-Arid environments-SEDiments), which has been developed within the joint Spanish-Brazilian-German research project SESAM (Sediment Export from Semi-Arid Catchments: Measurement and Modelling) and utilized to simulate North-East Brazil’s conditions \cite{Guntner2002,Guntner2004,Mueller2010}. The model simulates the runoff and erosion processes at meso-scale, for domains ranging from several hundreds to thousands of square kilometers. It uses a hierarchical top-down disaggregation scheme, dividing each sub-basin of the model into landscape units (based on the Soil and Terrain Digital Database (SOTER) concept \cite{Oldeman1993}), represented by multiple terrain components (defined by slope-gradient, length, soil and soil-vegetation components). The hydrological module is fully described in \citeA{Guntner2002} and \citeA{Guntner2004}, implementing for example equations to account for interception losses, evaporation and transpiration (modified Penman-Monteith approach \cite{Shuttleworth1985}) and for infiltration (Green-Ampt approach \cite{Green1911}) and other soil and vegetation related processes. The model can simulate both medium and large  centralized strategic reservoirs and small diffuse networks of reservoirs \cite{Guntner2004}. The user manual, the link to the source code and other useful sources can be found at \url{github.com/TillF/WASA-SED}. Detailed WASA-SED model parameterization for Banabuiú river basin was carried out by \citeA{Costa2013}, and adopted in this work.

\subsubsection{Small reservoirs representation}
The modeling approach to the reservoirs needs to classify them into strategic and small reservoirs, according to location and size \cite{Guntner2004}. The strategic reservoirs listed in Table~\ref{tab1} are individually parameterized in WASA-SED, also providing their daily regulated discharge time series. Their water balance is calculated explicitly and individually for each reservoir. Due to the uncontrolled and unmonitored nature of small reservoirs composing the DRN, not enough information is available concerning dam building, location, size and water use. They are grouped into 5 storage capacity classes and the water balance is computed for a hypothetical reservoir with mean characteristics, representative of each class and sub-basin. The water storage volume is then given by the product of the water storage of the representative reservoir by the total number of reservoirs from that class within a sub-basin. The generated runoff of the sub-basin is distributed among the reservoir classes through a cascade routing scheme starting from the lowest class to the highest, using a weighting factor computed as a ratio between the runoff contributing area of that reservoir class and the sub-basin area \cite{Mamede2018}. A detailed description of the calculation of water fluxes through the classes of reservoirs can be found in \citeA{Mamede2008}, while equations and other information about the hydrological and reservoir module can be found in \citeA{Guntner2002}.

\subsubsection{Calibration and validation}
Calibration was focused on the “scaling factor” parameter, which modifies the soil saturated hydraulic conductivity, which is the most sensitive parameter \cite{Guntner2002,WASA-SEDManual2021}. The calibration was carried out automatically following the sub-basins routing path of the model from upstream to downstream, testing a set of 35 values on each sub-basin (ranging from 0.2 to 7). For each run, five model performance indexes were computed (R2, NSE, KGE, PBIAS and NRMSE) comparing the modeled volume time series for the sub-basin’s strategic reservoir (or the direct downstream one in case the sub-basin did not present one itself) with the observed one. The performance indexes were selected based on recent usage in hydrological modeling \cite{Marahatta2021,Knoben2019,Uniyal2019}. The parameter that would produce the best performance was selected. The model was calibrated with 70\% of the series, running it from 1980 to 2006, it was then validated on the remaining 30\% of the series, from 2007 to 2018. Wet, normal and dry years were well represented in both the selected periods (calibration and validation) of the gauges. Therefore, the high inter-annual streamflow variability (CV greater than 1) was taken into account for both calibration and validation periods.

\subsubsection{Scenarios generated and used in the analyses}\label{sec:scenarios}

%Figure 2
\begin{figure}
 \noindent\includegraphics[width=\textwidth]{images/Figure_2.png}
 \caption{Visualization of the modeled scenarios design: Large Reservoirs (LR), centralized reservoirs only; All Reservoirs (AR), centralized reservoirs and DRN; Naturalized (N), Arrojado Lisboa only; and Small Reservoirs (SR), DRN and Arrojado Lisboa only.}
 \label{fig2}
\end{figure}

Four different configurations were modeled, as visualized in Figure~\ref{fig2}. The All Reservoirs (AR) configuration represents the real condition, with the co-existence of the small and large reservoirs. The dense reservoir network was removed to keep only large reservoirs in the Large Reservoirs (LR) scenario. These were also removed to obtain two scenarios not influenced by them: Small Reservoirs (SR), representing only small reservoirs and Arrojado Lisboa, and Naturalized (N), representing only Arrojado Lisboa. The only reservoir present in all the scenarios is Arrojado Lisboa, the most downstream, used as a reference for the comparisons in all the analyses. The differences between AR and LR scenarios have been explored in order to assess the effects of the DRN on droughts’ evolution in time and space. N and SR scenarios have been useful to explore the DRN effects without the influence of large reservoirs.

\subsection{Drought analyses}\label{sec:analyses}
To effectively consider the effects that the existence of a DRN produces on drought evolution, two domains have been considered: time, to address impacts on frequency and duration of drought, and space, to include the effects on water distribution. This dual aspect has been analyzed through the Drought Cycle Analysis (\ref{sec:DCA}) \cite{RibeiroNeto2022} and the Downstreamness Analysis (\ref{sec:dwn}) \cite{VanOel2011}. The different scenarios created through the WASA-SED model will be compared with these methods in order to estimate the influence of the DRN on hydrological drought evolution and propagation.

\subsubsection{Drought Cycle Analysis}\label{sec:DCA}

%Figure 3
\begin{figure}
 \noindent\includegraphics[width=\textwidth]{images/Figure_3.jpg}
 \caption{Graphical representation of the Drought Cycle Analysis, from \citeA{RibeiroNeto2022}. WSI: Water Storage Index, PI: Precipitation Index.}
 \label{fig3}
\end{figure}

The Drought Cycle Analysis (DCA) is a recent method proposed by \citeA{RibeiroNeto2022}, based on a combination of precipitation and hydrological indexes. It classifies drought events into four possible stages. The indexes values are positioned in quadrants in the coordinate system shown in Figure~\ref{fig3}, with a hue and tone variation scheme to represent the intensity of drought event as well as its classification, respectively.\\
The first quadrant is the non-occurrence of drought: positive values of both Precipitation Index (PI, vertical axis) and Water Storage Index (WSI, horizontal axis). The next quadrant indicates meteorological drought, when there is a precipitation deficit (PI < 0) but it has not yet affected water storage (WSI > 0). In the third quadrant, hydro-meteorological drought takes place, which is the coexistence of meteorological and hydrological drought. Water storage is affected by the persistence of meteorological drought and reservoirs deplete consistently. Therefore, both indexes reach negative values. The final quadrant is the "Hydrological drought quadrant", which is characterized by the persistence of hydrological drought (WSI < 0) after the end of meteorological drought (PI > 0).\\
Standardized indexes are used as they allow a better division in the quadrants proposed: negative values of the index mean a scarcity condition, while positive values mean the opposite. In order to compute the analysis, the Precipitation Index and the Water Storage Index can be chosen from the many available. The same procedure as \citeA{RibeiroNeto2022} has been followed. The Standardized Precipitation Index (SPI) over 12 months (SPI-12) has been utilized as PI (WMO, 2012). The Volume Deviation (VD) of Arrojado Lisboa reservoir (the largest and most downstream one) is utilized as a proxy for the water storage of the study area (WSI). VD is an index considering the deviation of a reservoir’s volume from the half of its total capacity. It can thus vary from -1 to 1, with positive values meaning a volume higher than half of the total capacity and negative values meaning the opposite. The indexes are computed at monthly resolution, the results of the Drought Cycle Analysis will then have a monthly resolution themselves. This method considers a meteorological index for monitoring droughts (the vertical axis of the drought wheel) and information directly related to the impact that this can cause (the horizontal axis of the drought wheel). This allows us to identify drought events defined as disasters related to the exceptional lack of water that is prejudicial for human activities or environmental demands. Further details and discussion on this method can be found in \citeA{RibeiroNeto2022}.

\subsubsection{Downstreamness analysis}\label{sec:dwn}
The downstreamness concept aims to analyze the availability and distribution of water resources in a river basin, first introduced in \citeA{VanOel2009} and successively developed and used to analyze basin closure and to diagnose hydrological droughts \cite{VanOel2011,VanOel2018,HvanLangen2021}. The downstreamness of a location ($D_{x}$), for example a reservoir’s dam outlet, is the ratio of its upstream catchment area ($A_{up}$) to the entire river basin area ($A_{tot}$) (Equation 2). Higher the index, the more downstream the location x will be.\\
